---
editor: 
  markdown: 
    wrap: 72
---

EDA

```{r}
library(tidyverse)
library(ggplot2)
```

The data set used in this project is called the "**Diabetes Health
Indicators Dataset**". Diabetes is a large problem in America; as of
2018 34.2 million Americans have diabetes and 88 million have
prediabetes. The original data set from the Behavioral Risk Factor
Surveillance System (BRFSS) contains responses from 441,455 and has 330
features.

This project looks at a few variables in particular:

Smoker: Have you smoked at least 100 cigarettes in your entire life (YES
OR NO)

PhysActivity: Physical activity in the last 30 days (YES OR NO)

Education: Education level

Age: Age category of subject

Sex: Male or Female

HeartDiseaseorAttack: Coronary heart disease (CHD) or myocardial
infarction (MI) (YES OR NO)

HighBP: No high BP or high BP BMI: BMI of subject

The goal of Exploratory Data Analysis (EDA) is to analyze and
investigate data sets to understand and summarize the data before
applying them to models. This helps validate results from models and
other applications.

**EDA Process:** Understand how data is stored Data validation Determine
rate of missing values Clean data up Investigate distributions Apply
transformations

```{r}
diabetes <- read.csv("diabetes_binary_health_indicators_BRFSS2015.csv")
```

Cleaning the data and turning the variables into factors.

```{r}
diabetes$Diabetes_binary <- factor(diabetes$Diabetes_binary, levels = c(0, 1, 2), labels = c("No diabetes", "prediabetes", "diabetes"))
diabetes$HighBP <- factor(diabetes$HighBP, levels = c(0, 1), labels = c("No high BP", "High BP"))
diabetes$HighChol <- factor(diabetes$HighChol, levels = c(0, 1), labels = c("No high cholesterol", "High cholesterol"))
diabetes$CholCheck <- factor(diabetes$CholCheck, levels = c(0, 1), labels = c("No chol check in 5 years", "Yes chol check in 5 years"))
diabetes$Smoker <- factor(diabetes$Smoker, levels = c(0, 1), labels = c("No", "Yes"))
diabetes$Stroke <- factor(diabetes$Stroke, levels = c(0, 1), labels = c("No", "Yes"))
diabetes$HeartDiseaseorAttack <- factor(diabetes$HeartDiseaseorAttack, levels = c(0, 1), labels = c("No", "Yes"))
diabetes$PhysActivity <- factor(diabetes$PhysActivity, levels = c(0, 1), labels = c("No", "Yes"))
diabetes$Fruits <- factor(diabetes$Fruits, levels = c(0, 1), labels = c("No", "Yes"))
diabetes$Veggies <- factor(diabetes$Veggies, levels = c(0, 1), labels = c("No", "Yes"))
diabetes$HvyAlcoholConsump <- factor(diabetes$HvyAlcoholConsump, levels = c(0, 1), labels = c("No", "Yes"))
diabetes$AnyHealthcare <- factor(diabetes$AnyHealthcare, levels = c(0, 1), labels = c("No", "Yes"))
diabetes$NoDocbcCost <- factor(diabetes$NoDocbcCost, levels = c(0, 1), labels = c("No", "Yes"))
diabetes$GenHlth <- factor(diabetes$GenHlth, levels = c(1,2,3,4,5), labels = c("Excellent", "Very good", "Good", "Fair", "Poor"))
diabetes$DiffWalk <- factor(diabetes$DiffWalk, levels = c(0, 1), labels = c("No", "Yes"))
diabetes$Sex <- factor(diabetes$Sex, levels = c(0, 1), labels = c("Female", "Male"))
diabetes$Age <- factor(diabetes$Age, levels = c(1,2,3,4,5,6,7,8,9,10,11,12,13), labels = c("Age 18 to 24","Age 25 to 29","Age 30 to 34","Age 35 to 39","Age 40 to 44","Age 45 to 49","Age 50 to 54","Age 55 to 59","Age 60 to 64","Age 65 to 69","Age 70 to 74","Age 75 to 79","Age 80 or older"))
diabetes$Education <- factor(diabetes$Education, levels = c(1,2,3,4,5,6,9), labels = c("Never attended school or only kindergarten","Grades 1 through 8 (Elementary)", "Grades 9 through 11 (Some high school)", "Grade 12 or GED (High school graduate)", "College 1 year to 3 years (Some college or technical school)", "College 4 years or more (College graduate)", "Refused"))
diabetes$Income <- factor(diabetes$Income, levels = c(1,2,3,4,5,6,7,8,77,99), labels = c("Less than $10,000", "Less than $15,000 ($10,000 to less than $15,000)", "Less than $20,000 ($15,000 to less than $20,000)", "Less than $25,000 ($20,000 to less than $25,000)", "Less than $35,000 ($25,000 to less than $35,000)", "Less than $50,000 ($35,000 to less than $50,000)", "Less than $75,000 ($50,000 to less than $75,000)", "$75,000 or more", "Don’t know/Not sure", "Refused"))
diabetes$BMI <- format(diabetes$BMI, nsmall = 1)
diabetes$MentHlth <- format(diabetes$MentHlth, nsmall = 1)
diabetes$PhysHlth <- format(diabetes$PhysHlth, nsmall = 1)
diabetes$BMI <- as.double(diabetes$BMI)
```

Checking for missing values in the data.

```{r}
colSums(is.na(diabetes))
```

Looking at the structure of the data.

```{r}
str(diabetes)
```

Summarizing the data set.

```{r}
summary(diabetes)
```

Creating contingency tables.

```{r}
smoker_table <- table(diabetes$Smoker, diabetes$HeartDiseaseorAttack)
rownames(smoker_table) <- c("Not smoked 100 cigarettes", "Smoked 100 cigarettes")
colnames(smoker_table) <- c("No CHS/MI", "Yes CHS/MI")
smoker_table_margins <- addmargins(smoker_table)
smoker_table_margins
```

```{r}
table2 <- table(diabetes$Age, diabetes$PhysActivity, diabetes$HighBP)
colnames(table2) <- c("No Phys Activity in last 30 days", "Physical activity in last 30 days")
table2
```

```{r}
typeof(diabetes$BMI)
```

```{r}
diabetes |>
   summarize(mean_bmi = mean(BMI),
            median_bmi = median(BMI))
```
From the plot below you can see that age 60-64 has the largest amount of people in the data set. The smallest group of ages included in the data set is 18-24, which is interesting.
```{r}
plot1 <- ggplot(diabetes, aes(x=Age)) + 
  geom_bar(fill = "skyblue") + 
  labs(title="Bar Plot of Age", x="Age", y="Count") +
theme(axis.text.x = element_text(angle = 90, hjust = 1))
plot1
```

The boxplot below looks at BMI vs. Sex. Men have a slightly higher median when compared to women. women have a larger distribution of BMI.
```{r}
boxplot1 <- ggplot(diabetes, aes(x = Sex, y = BMI)) +
  geom_boxplot(fill = "skyblue") +
  labs(title="Box Plot of BMI by Sex", x="Sex", y="BMI")
boxplot1
```
From the graph below we can see that there are way more people who have not had heart disease or a heart attack. There is around 25,000 people who have had a heart attack and are included in the data set.
```{r}
plot2 <- ggplot(diabetes, aes(x=HeartDiseaseorAttack)) + 
  geom_bar(fill = "skyblue") + 
  labs(title="Bar Plot of Heart Disease or Attack", x="Heart Disease or Attack?", y="Count") +
theme(axis.text.x = element_text(angle = 90, hjust = 1))
plot2
```

[Click here for the Modeling Page](Modeling.html)
